Whole-Body Manipulation Planning for Humanoid Robots

### 2012-06-15 paper.tex
Fix minor typos, modify some mathematical terms.

When  planning a  whole-body motion  for a  humanoid robot, one difficult
challenge is to cope with  the curse of dimensionality. The complexity
of   motion  planning  is   exponential  in   the  dimension   of  the
configuration  space ($\mathcal{C}$)  to explore.  When  dealing with
high-dimensional configuration  spaces, it is  typically impossible to
explicitly represent  them, leading to the use  of randomized sampling
techniques  to solve  global planning  problems. In  the  past fifteen
years,  \textit{Probabilistic Roadmaps} \cite{kavraki1996prp} and  
\textit{Rapidly exploring Random  Trees} (RRT) 
\cite{kuffner00rrtconnect}  have been  developed and  used to  solve many
high-dimensional   planning  problems.
When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
difficulty  is to  take into  account stability  constraints,  i.e. to
generate  random  configurations   on  zero  volume  submanifolds  of
$\mathcal{C}$. This problem has been investigated with success during
the last few years, \cite{Berenson15032011} presents a exhaustive survey
of jacobian-based methods. Other recent contributions \cite{porta2012randomized}
present sophisticated constrained planning techniques based on higher-dimensional
continuation. Section \ref{sec:wb} presents a simple adaptation
of the RRT algorithm to constrained motion planning, that was first
introduced in \cite{dalibard09}.

@@

This section presents an algorithm for constrained motion planning. 
The problem solved here
differs from classic configuration space approaches in two ways:
\begin{enumerate}
\item the set of valid configurations is defined implicitly, as the ones satisfying a given 
  set of inverse kinematic tasks;
\item the goal is also defined implicitely, by another set of inverse kinematic tasks.
\end{enumerate}
In this presentation, we use a prioritized inverse kinematic solver, or local planner,
 as a black box. We will not describe the details of its implementation. They are presented in 
\cite{yoshida2006tds}, and are out of scope of the present work.

During global planning, we will consider several types of
tasks and constraints for various reasons:
\begin{itemize}
\item  Static stability: the center of mass of the robot stays at
  the vertical of the support polygon center, the two feet have fixed position and
  orientation.
\item End-effector position and orientation: the goals of some problems presented
  in the experimental section of this paper are defined as robot hand configuration,
  or gaze direction.
\item Configuration task: our adaptation of randomized motion planning algorithms 
  uses tasks defined as the distance towards a given configuration in $\mathcal{C}$. 
  This will be detailed in the following of the section.
\end{itemize}

@@

The motion presented in Fig.~\ref{fig:wb-shelves} looks very complicated.
The reasons are that the environment is complex and HRP-2 kinematic chain is highly constrained.
The motion has been optimized based on these constraints. Note that the
accessibility space of this robot is not wide enough to accomplish this task further 
away from the shelves, which could have eased the motion. One possibility to extend the
robot accessibility space is to automatically change the support polygon to solve the task.
This is an ongoing research field, \cite{OussamaKanoun06102010} proposes some 
local solutions. This has not been integrated in our global planning algorithm yet.

@@

This section presents a \textit{constructive} proof that any such statically stable, 
collision-free path for a legged robot sliding
on the ground can be approximated by a collision-free dynamic walk trajectory.
The proof is based on ideas from control theory, in particular small-space 
controllability. It also uses the fact that balanced criteria for dynamic walk are different
from the ones for static stability. An important point is that our result and method do not hold
for a biped robot that walks in a quasi-static way.

@@

A  robotic system  is  controllable  if  for any two  configurations
$q_1$ and $q_2$,  there exists  a
trajectory  going  from  $q_1$ to  $q_2$.  It  is  
\textit{small-space  controllable} if for  any configuration  $q$, 
for any $\epsilon >0$, there
exist $\eta >0$ such that all the configurations contained in the ball of center
$q$ and radius $\eta$ are reachable by trajectories included in the
ball of center $q$ and radius $\epsilon$. Fig.~\ref{fig:ssc1} shows an illustration
of this property.

@@

\begin{proof}[Proof of Theorem \ref{thm:ssc}]
  Let $\mathcal{C}$ be the configuration space of a small-space controllable robot, and 
  $\mathcal{C}_{free} \subset \mathcal{C}$ the set of collision-free configurations. We
  consider in-contact configurations as colliding, so $\mathcal{C}_{free}$ is an open set.
  Let $\tau : [0,1] \rightarrow \mathcal{C}_{free}$ be a collision-free path. For every $x \in [0,1]$,
  $\tau(x) \in \mathcal{C}_{free}$, so there exists $\epsilon_x$ such that the open ball 
  $B(\tau(x),\epsilon_x)$ of center $\tau(x)$ and radius $\epsilon_x$ is included in 
  $\mathcal{C}_{free}$. The small-space controllability property states that for every $x$,
  there exist $\eta_x > 0$ such that every configuration $q \in B(\tau(x),\eta_x)$ is reachable 
  from $\tau(x)$ by a trajectory included in $B(\tau(x),\epsilon_x)$.

@@

\subsubsection{Small-Time \textit{versus} Small-Space Controllability}
In the control theory literature, the property used is usually \textit{small-time controllability}, 
which states that for  any configuration  $q$, for  any time
$T>0$, the set of configurations accessible from $q$ in time less than
$T$ forms a  neighborhood of $q$. When accelerations and velocities are bounded,
small-time controllability implies small-space controllability, which is why 
a lot of motion planning previous work only refers to the sufficient small-time controllability
property. However, the reciprocate is not necessarily true:  a system can be 
small-space controllable and
not small-time, if the trajectories generated by its controller are arbitrarily long.
The important property
in terms of motion planning application is the small-space controllability, as 
Theorem~\ref{thm:ssc} shows. In the following, we show that legged robots
are small-space controllable, but  not that they are small-time controllable.
In fact, the  control method that we present does not follow the small-time controllability
property. For the sake of clarity, we have chosen to make the distinction between these
two controllability properties.

@@

This section  discusses the small-space controllability of walking robots. To clarify
the presentation, we consider a simplified model of legged robot consisting of two feet
of null mass and a punctual mass free to move in three dimensions.
We do not consider 
the kinematic chains between the feet and the mass. The robot is walking on a flat terrain,
and the feet are assumed to have a positive surface. For our presentation, it is not 
necessary to consider the feet height, so the configuration space of the robot is:
\[
\mathcal{C} = SE(2) \times SE(2) \times \mathbb{R}^3
\]
It is of dimension 9.

The balanced walking conditions for a quasi-static walking robot are that the punctual
mass, or CoM, should always be over the support polygon (the convex hull of the two feet), and 
one foot can move iff the CoM is over the other foot. Similarly, the walking
conditions for a dynamic walking robot are that the ZMP should
always be in the robot support polygon, and on foot can move iff the ZMP is over the other 
foot. For a precise description of dynamically balanced walking conditions, the reader can refer
to \cite{wieber2002}. Under these assumptions, the following result holds:

@@

The first claim is straightforward. Let the robot be in a configuration $q$ where the two
feet are separated from a positive distance. Let  $L>0$ be the positive horizontal distance
between the CoM and the left foot (if the CoM is over the left foot, we can consider similarly 
the right foot). For any $\epsilon < L$, any valid trajectory starting from $q$, included in
the ball of center $q$ and radius $\epsilon$, is such that the CoM is never over the left foot.
Given the quasi-static walk conditions, the right foot of the robot is fixed along 
the trajectory. Thus, the set of accessible configurations from $q$ by staying inside 
$B(q,\epsilon)$ does not form a neighborhood of $q$, since it does not include any configuration
where the right foot has moved. This shows that the robot is not small-space controllable.

@@

Let us now consider a dynamic walking robot. Starting from any valid static configuration, 
the CoM can move vertically without affecting balance, so there is no need to 
consider this degree of freedom in the following. If the CoM is not over the edge of the support
polygon, it is possible to move it in a quasi-static way inside a neighborhood of its current position that projects itself over the support polygon.
It is thus sufficient and necessary to prove that for any $\epsilon >0$, it is possible to move
the feet while keeping the CoM inside a neighborhood of size $\epsilon$. Let such $\epsilon >0$
be arbitrarily fixed.

>>>>

When  planning a  whole-body motion  for a  humanoid robot, one difficult
challenge is to cope with  the curse of dimensionality. The complexity
of   motion  planning  is   exponential  in   the  dimension   of  the
configuration  space ($\mathcal{C}$)  to explore.  When  dealing with
high-dimensional configuration  spaces, it is  typically impossible to
explicitly represent  them, leading to the use  of randomized sampling
techniques  to solve  global planning  problems. In  the  past fifteen
years,  \textit{Probabilistic Roadmaps} \cite{kavraki1996prp} and  
\textit{Rapidly exploring Random  Trees} (RRT) 
\cite{kuffner00rrtconnect}  have been  developed and  used to  solve many
high-dimensional   planning  problems.
When  using  sampling  techniques  on  a humanoid  robot,  another difficulty
is to  take into  account stability  constraints,  i.e. to
generate  random  configurations   on  zero  volume  submanifolds  of
$\mathcal{C}$. This problem has been investigated with success during
the last few years, \cite{Berenson15032011} presents a exhaustive survey
of jacobian-based methods. Other recent contributions \cite{porta2012randomized}
present sophisticated constrained planning techniques based on higher-dimensional
continuation. Section \ref{sec:wb} presents a simple adaptation
of the RRT algorithm to constrained motion planning, that was first
introduced in \cite{dalibard09}.

@@

This section presents an algorithm for constrained motion planning. 
The problem solved here
differs from classic configuration space approaches in two ways:
\begin{enumerate}
\item the set of valid configurations is defined implicitly, as the ones satisfying a given 
  set of inverse kinematic tasks;
\item the goal is also defined implicitly, by another set of inverse kinematic tasks.
\end{enumerate}
In this presentation, we use a prioritized inverse kinematic solver, or local planner,
 as a black box. We will not describe the details of its implementation. They are presented in 
\cite{yoshida2006tds}, and are out of scope of the present work.

During global planning, we will consider several types of
tasks and constraints for various reasons:
\begin{itemize}
\item  Static stability: the center of mass of the robot stays at
  the vertical of the support polygon center, the two feet have fixed position and
  orientation.
\item End-effector position and orientation: the goals of some problems presented
  in the experimental section of this paper are defined as robot hand configuration,
  or gaze direction.
\item Configuration task: our adaptation of randomized motion planning algorithms 
  uses tasks defined as the distance towards a given configuration in $\mathcal{C}$. 
  This will be detailed in the following section.
\end{itemize}

@@

The motion presented in Fig.~\ref{fig:wb-shelves} looks very complicated.
The reasons are that the environment is complex and HRP-2 kinematic tree is highly constrained.
The motion has been optimized based on these constraints. Note that the
accessibility space of this robot is not wide enough to accomplish this task further 
away from the shelves, which could have eased the motion. One possibility to extend the
robot accessibility space is to automatically change the support polygon durint the motion to solve the task.
This is an ongoing research field, \cite{OussamaKanoun06102010} proposes some 
local solutions. This has not been integrated in our global planning algorithm yet.

@@

This section presents a \textit{constructive} proof that any such statically stable, 
collision-free path for a legged robot sliding
on the ground can be approximated by a collision-free dynamic walk trajectory.
The proof is based on ideas from control theory, in particular small-space 
controllability. It also uses the fact that balance criteria for dynamic walk are different
from the ones for static stability. An important point is that our result and method do not hold
for a biped robot that walks in a quasi-static way.

@@

A  robotic system  is  controllable  if  for any two  configurations
$q_1$ and $q_2$,  there exists  a
trajectory  going  from  $q_1$ to  $q_2$.  It  is  
\textit{small-space  controllable} if for all configurations  $q$, 
for all $\epsilon >0$, there
exists $\eta >0$ such that all the configurations contained in the ball of center
$q$ and radius $\eta$ are reachable by trajectories included in the
ball of center $q$ and radius $\epsilon$. Fig.~\ref{fig:ssc1} shows an illustration
of this property.

@@

\begin{proof}[Proof of Theorem \ref{thm:ssc}]
  Let $\mathcal{C}$ be the configuration space of a small-space controllable robot, and 
  $\mathcal{C}_{free} \subset \mathcal{C}$ the set of collision-free configurations. We
  consider in-contact configurations as colliding, so $\mathcal{C}_{free}$ is an open set.
  Let $\tau : [0,1] \rightarrow \mathcal{C}_{free}$ be a collision-free path. For all $x \in [0,1]$,
  $\tau(x) \in \mathcal{C}_{free}$, so there exists $\epsilon_x$ such that the open ball 
  $B(\tau(x),\epsilon_x)$ of center $\tau(x)$ and radius $\epsilon_x$ is included in 
  $\mathcal{C}_{free}$. The small-space controllability property states that for all $x$,
  there exists $\eta_x > 0$ such that every configuration $q \in B(\tau(x),\eta_x)$ is reachable 
  from $\tau(x)$ by a trajectory included in $B(\tau(x),\epsilon_x)$.

@@

\subsubsection{Small-Time \textit{versus} Small-Space Controllability}
In the control theory literature, the property used is usually \textit{small-time controllability}, 
which states that for  all configurations  $q$, for  all times
$T>0$, the set of configurations accessible from $q$ in time less than
$T$ forms a  neighborhood of $q$. When accelerations and velocities are bounded,
small-time controllability implies small-space controllability, which is why 
a lot of motion planning previous work only refers to the sufficient small-time controllability
property. However, the reciprocate is not necessarily true:  a system can be 
small-space controllable and
not small-time, if the trajectories generated by its controller are arbitrarily long.
The important property
in terms of motion planning application is the small-space controllability, as 
Theorem~\ref{thm:ssc} shows. In the following, we show that legged robots
are small-space controllable, but  not that they are small-time controllable.
In fact, the  control method that we present does not follow the small-time controllability
property. For the sake of clarity, we have chosen to make the distinction between these
two controllability properties.

@@

This section  discusses the small-space controllability of walking robots. To clarify
the presentation, we consider a simplified model of a legged robot consisting of two feet
of null mass and a punctual mass free to move in three dimensions.
We do not consider 
the kinematic chains between the feet and the mass. The robot is walking on a flat terrain,
and the feet are assumed to have a positive surface. For our presentation, it is not 
necessary to consider the feet height, so the configuration space of the robot is:
\[
\mathcal{C} = SE(2) \times SE(2) \times \mathbb{R}^3
\]
It is of dimension 9.

The balanced walking conditions for a quasi-static walking robot are that the punctual
mass, or CoM, should always be over the support polygon (the convex hull of the two feet), and 
one foot can move iff the CoM is over the other foot. Similarly, the walking
conditions for a dynamic walking robot are that the ZMP should
always be in the robot support polygon, and one foot can move iff the ZMP is over the other 
foot. For a precise description of dynamically balanced walking conditions, the reader can refer
to \cite{wieber2002}. Under these assumptions, the following result holds:

@@

The first claim is straightforward. Let the robot be in a configuration $q$ where the two
feet are separated by a positive distance. Let  $L>0$ be the positive horizontal distance
between the CoM and the left foot (if the CoM is over the left foot, we can consider similarly 
the right foot). For all $\epsilon < L$, any valid trajectory starting from $q$, included in
the ball of center $q$ and radius $\epsilon$, is such that the CoM is never over the left foot.
Given the quasi-static walk conditions, the right foot of the robot is fixed along 
the trajectory. Thus, the set of accessible configurations from $q$ by staying inside 
$B(q,\epsilon)$ does not form a neighborhood of $q$, since it does not include any configuration
where the right foot has moved. This shows that the robot is not small-space controllable.

@@

Let us now consider a dynamic walking robot. Starting from any valid static configuration, 
the CoM can move vertically without affecting balance, so there is no need to 
consider this degree of freedom in the following. If the CoM is not over the edge of the support
polygon, it is possible to move it in a quasi-static way inside a neighborhood of its current position that projects itself over the support polygon.
It is thus sufficient and necessary to prove that for all $\epsilon >0$, it is possible to move
the feet while keeping the CoM inside a neighborhood of size $\epsilon$. Let such $\epsilon >0$
be arbitrarily fixed.
